\section{Models}
\label{models}

[[[TO DO]]]

This section defines the notion of model of a context.
%(recall that specs are contexts without variable and type variable declarations)

A model of a context is a mapping from names declared in the context to
suitable set-theoretic entities. For instance, a type name $\tnam$ of arity
$n$ is mapped to an $n$-ary function over sets (if $n=0$, the model maps the
type name simply to a set). The mapping is extended to all well-formed types,
which are mapped to sets, and to all well-typed expressions, which are mapped
to elements of the sets that their types map to. The model must satisfy all
the type definitions, op definitions, and axioms of the context.

It should be possible to prove the soundness of the rules to derive judgements
with respect to models.

Since higher-order logic is notoriously incomplete, it is not possible to
prove completeness of the rules to derive assertions. However, it should be
possible to prove completeness with respect to general (a.k.a.\ Henkin)
models. A general model is one in which the type $\tarrO$ is a subset of all
functions from $\typ_1$ to $\typ_2$, and not necessarily the set of all such
functions (as in standard models). Since there are more general models than
standard models (a standard model is also a general model but not all general
models are standard models), fewer formulas are true in all general models
than in all standard models.

Perhaps this section should also contain a proof of the consistency of the
\MS\ logic, analogously to the proof of the consistency of the higher-order
logic defined in \cite{andrews}.
